---
title: "The Area of Effect Concept"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Area of Effect Concept}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "svg",
  fig.ext = "svg",
  fig.width = 7,
  fig.height = 5
)
```

## The Problem: Border Truncation

When analyzing spatial data within political or administrative boundaries, a fundamental assumption is often violated: that the sampling region represents the ecological extent of the processes being studied.

Consider sampling species occurrences within a country. Observations near the border are influenced by conditions *outside* that country. A forest that spans the border, a river that crosses it, or simply the continuous nature of climate and habitat means that truncating at the border introduces systematic bias.

This is **border truncation**: the artificial constraint of ecological processes to administrative boundaries.

## What AoE Is Not

Before explaining what AoE does, it's important to clarify what it is *not*:
- **Not a buffer**: Buffers add a fixed distance. AoE scales proportionally from a reference point.
- **Not a distance-based operation**: There is no distance threshold or decay function.
- **Not a tunable parameter**: Scale is fixed at 1. This is a method, not a knob.

## The Area of Effect

The **area of effect** (AoE) is the spatial extent over which observations within a support are influenced by external conditions. It is computed by scaling the support geometry outward from a reference point (typically the centroid).

With scale fixed at 1, the transformation is:

$$p' = r + 2(p - r)$$

where:
- $r$ is the reference point (centroid)
- $p$ is each vertex of the support boundary
- $p'$ is the transformed vertex

This doubles the distance from the reference point to each boundary vertex, effectively expanding the support to twice its spatial extent while preserving shape and orientation.

## Core and Halo Classification

Points within the AoE are classified into two categories:

1. **Core**: Points inside the original support. These are fully contained within the sampling region and represent "pure" observations unaffected by border effects.

2. **Halo**: Points outside the original support but inside the expanded AoE. These observations are influenced by conditions in the border zone and may require different treatment in analysis.

Points outside the AoE are **pruned** (removed). They are too distant to be meaningfully related to the support region.

```{r classification, echo=FALSE, fig.cap="Point classification by AoE. Core points (green) are inside the original support. Halo points (orange) are in the expanded region. Points outside the AoE are pruned."}
library(sf)
library(areaOfEffect)

# Create support
support <- st_as_sf(
  data.frame(id = 1),
  geometry = st_sfc(st_polygon(list(
    cbind(c(2, 8, 8, 2, 2), c(2, 2, 8, 8, 2))
  ))),
  crs = NA
)

# Create AoE manually for visualization
centroid <- c(5, 5)
support_coords <- cbind(c(2, 8, 8, 2, 2), c(2, 2, 8, 8, 2))
aoe_coords <- centroid + 2 * (support_coords - centroid)
aoe_poly <- st_polygon(list(aoe_coords))

# Create points
set.seed(42)
n_pts <- 50
pts_coords <- cbind(
  runif(n_pts, -2, 12),
  runif(n_pts, -2, 12)
)
pts <- st_as_sf(
  data.frame(id = 1:n_pts),
  geometry = st_sfc(lapply(1:n_pts, function(i) st_point(pts_coords[i, ]))),
  crs = NA
)

# Classify
in_support <- st_intersects(pts, support, sparse = FALSE)[, 1]
in_aoe <- st_intersects(pts, st_sfc(aoe_poly), sparse = FALSE)[, 1]

pts$class <- ifelse(in_support, "core",
                    ifelse(in_aoe, "halo", "pruned"))

# Plot
par(mar = c(2, 2, 2, 2))
plot(st_geometry(st_sfc(aoe_poly)), border = "gray70", lty = 2,
     xlim = c(-3, 13), ylim = c(-3, 13), asp = 1)
plot(st_geometry(support), border = "black", lwd = 2, add = TRUE)
points(pts_coords[pts$class == "pruned", ], col = "gray60", pch = 4, cex = 0.8)
points(pts_coords[pts$class == "halo", ], col = "darkorange", pch = 16)
points(pts_coords[pts$class == "core", ], col = "forestgreen", pch = 16)
points(5, 5, pch = 3, cex = 1.5, lwd = 2)  # centroid
legend("topright",
       legend = c("Core", "Halo", "Pruned", "Centroid", "Original", "AoE"),
       col = c("forestgreen", "darkorange", "gray60", "black", "black", "gray70"),
       pch = c(16, 16, 4, 3, NA, NA),
       lty = c(NA, NA, NA, NA, 1, 2),
       lwd = c(NA, NA, NA, 2, 2, 1),
       bg = "white")
```

## The Scale Parameter

The `scale` parameter controls how far the AoE extends beyond the original support. With scale `s`, the linear multiplier is `1 + s`, and the area multiplier is `(1 + s)²`.

Two values have special geometric meaning:

- **`sqrt(2) - 1` ≈ 0.414** (default): Equal core and halo areas
- **`1`**: Equal linear distance inside and outside the boundary

## Why Equal Area is the Default

The default scale produces equal core and halo areas. This is not arbitrary—it reflects a philosophical position about what "influence" means.

### The Symmetry Argument

When we say a point in the halo is "influenced by" the support region, we're making a claim about spatial relevance. The question is: how much relevance should we grant to the outside?

Equal area says: **the outside matters as much as the inside**.

This is the maximally symmetric choice. Any other ratio implies that either:
- The core is more important than the halo (halo smaller), or
- External conditions dominate internal ones (halo larger)

Without domain-specific knowledge to justify asymmetry, equal weighting is the principled default.

### The Information-Theoretic View

Consider the AoE as defining a probability distribution over space: "where might conditions relevant to this support come from?"

Equal areas means equal prior probability mass inside and outside the original boundary. This is the maximum-entropy choice—it encodes no bias toward internal or external dominance.

Scale = 1 (the "one full stamp" expansion) gives a 1:3 ratio, implying the outside is three times more relevant than the inside. This may be appropriate for some analyses (e.g., highly mobile species), but it's a strong assumption to make by default.

### The Geometric Inevitability

The formula `s = √2 - 1` is not a tuned parameter. It's the *unique* solution to the constraint "core equals halo." This value emerges from geometry itself:

$$
(1 + s)^2 - 1 = 1 \implies s = \sqrt{2} - 1
$$

There's something satisfying about a default that isn't chosen but *derived*. It removes a degree of freedom from the analyst and replaces it with a principled constraint.

### When to Override

Use `scale = 1` when:
- You want the classic "one full stamp" expansion
- Your domain knowledge suggests external conditions dominate
- You're comparing with previous work that used this convention

Use custom scales when:
- You have empirical data on influence decay
- Sensitivity analysis requires exploring the parameter space
- Domain expertise justifies a specific ratio

## Hard vs Soft Boundaries

AoE distinguishes between two types of boundaries:

**Political borders (soft)**: Administrative lines have no ecological meaning. The AoE freely crosses them. A country border does not stop species from dispersing or climate from varying.

**Sea boundaries (hard)**: Physical barriers like coastlines are true boundaries. The optional `mask` argument enforces these constraints by intersecting the AoE with a land polygon.

```{r mask-concept, echo=FALSE, fig.cap="Hard boundaries constrain the AoE. The dashed line shows the theoretical AoE; the solid gray area shows the AoE after applying a land mask."}
# Create a coastal support
support_coastal <- st_as_sf(
  data.frame(id = 1),
  geometry = st_sfc(st_polygon(list(
    cbind(c(4, 8, 8, 4, 4), c(2, 2, 6, 6, 2))
  ))),
  crs = NA
)

# Theoretical AoE
centroid_c <- c(6, 4)
coastal_coords <- cbind(c(4, 8, 8, 4, 4), c(2, 2, 6, 6, 2))
aoe_coastal_coords <- centroid_c + 2 * (coastal_coords - centroid_c)
aoe_coastal <- st_polygon(list(aoe_coastal_coords))

# Land mask (irregular coastline)
land <- st_polygon(list(cbind(
  c(0, 10, 10, 7, 5, 3, 0, 0),
  c(0, 0, 5, 6, 5.5, 7, 6, 0)
)))

# Masked AoE
aoe_masked <- st_intersection(aoe_coastal, land)

par(mar = c(2, 2, 2, 2))
plot(aoe_coastal, border = "gray70", lty = 2,
     xlim = c(-1, 11), ylim = c(-1, 9), asp = 1)
plot(land, col = NA, border = "steelblue", lwd = 2, add = TRUE)
plot(aoe_masked, col = rgb(0.5, 0.5, 0.5, 0.3), border = NA, add = TRUE)
plot(st_geometry(support_coastal), border = "black", lwd = 2, add = TRUE)
points(6, 4, pch = 3, cex = 1.5, lwd = 2)

# Add "SEA" label
text(8, 7, "SEA", col = "steelblue", font = 2, cex = 1.2)

legend("topleft",
       legend = c("Support", "AoE (theoretical)", "AoE (masked)", "Coastline"),
       col = c("black", "gray70", rgb(0.5, 0.5, 0.5, 0.5), "steelblue"),
       lty = c(1, 2, NA, 1),
       lwd = c(2, 1, NA, 2),
       pch = c(NA, NA, 15, NA),
       pt.cex = 2,
       bg = "white")
```

## Multiple Supports

Real-world analyses often involve multiple administrative regions (countries, provinces, protected areas). AoE handles these naturally:

- Each support is processed independently
- Each uses its own centroid as reference
- Points can fall within multiple AoEs (when regions are adjacent)
- Output is in long format: one row per point-support combination

This enables cross-border analyses and studies of nested administrative structures without repeated preprocessing.

## Summary

The area of effect provides a principled correction for border truncation in spatial analysis:

- **Principled default**: Scale = √2 − 1, giving equal core and halo areas
- **Geometric derivation**: The default emerges from symmetry, not tuning
- **Categorical output**: Core, halo, or pruned
- **Soft/hard boundaries**: Political borders ignored, physical barriers respected
- **Multiple supports**: Process many regions at once

The result is a reproducible, interpretable method that can be consistently applied across studies.
